\Large \bf Cyclone: Replication Middleware for NVM Applications

### 2016-09-18 paper/paper.tex
One more graph.
Prepare for short paper.

\documentclass[twocolumn]{article}
\usepackage{times}
\usepackage{parskip}
\begin{document}
\title{Cyclone: Replication Middleware for NVM Clusters}
\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
\maketitle
\begin{abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. This will allow a broad range of new applications that exploit the
combination of durability with performance far superior to traditional durable
storage media such as NAND flash. However, durability does not guaratee
availability and without a means to survive machine or storage medium failures
these applications cannot match the survivability offered by sophisticated
distributed file systems, databases and key value stores.

Cyclone is replication middleware aimed at helping application developers
targeting new memory types to easily add fault tolerance to their
applications. Cyclone eschews external co-ordination services like zookeeper and
instead replicates RPC calls from clients being servered out of durable memory
across a quorum of clusters using the RAFT consensus protocol. Cyclone also
provides a client side library to enable transparent failover. Finally, Cyclone
is designed to provide at most once semantics to RPC calls. We demonstrate in
this paper that this is sufficient to design applications that support
distributed transactions without any need for the developer to be concerned with
fault recovery - a challenging aspect of designing systems that support
distributed transaction and at the same time are fault tolerant.

Cyclone is also designed to provide maximum performance without being dependent
on high end networking stacks - such as RDMA/Infiniband combinations.  To enable
the lowest possible latency we use DPDK - a userspace networking stack and
designed our implementation of RAFT to use zero copy for the replication
process. To enable maximum throughput under load we extensively batch
replication operations and support multiple concurrent instances of RAFT to
maximize throughput. As a result, Cyclone can replicate as many as 4 million 100
byte RPC messages every second on a very modest commodity 10 GigE network. Under
low load, the latency seen by a client for an RPC call the requires replication
is of the order of 20 us.
\end{abstract}

\section{Introduction}
TBD
\section{Background: Single Node Programming Model}
Cyclone builds on top of Intel's Non Volatile Memory Library (NVML) and provides
transparent replication to programs written to use NVML for persistence on a
single node. We describe NVML's programming model in this section to provide
adequate background for the rest of this paper.

NVML provides special functions to allocate and free persistent memory as well
as fat pointers to persistent memory to ensure that referential integrity in the
heap is maintained across restarts - since the virtual memory ranges into which
persistent memory is mapped may change.

\begin{figure}
{\scriptsize
\begin{verbatim}

struct ll_node {
  int value;
  TOID(ll_node) next;
};

void insert_after(TOID(ll_node) prev, 
                  TOID(ll_node) new_node)
{
  D_RW(new_node)->next = D_RO(prev)->next;
  D_RW(prev)->next = new_node;
}

\end{verbatim}
}
\caption{A Persistent Linked List of Integers}
\label{fig:example}
\end{figure}

Figure~\ref{fig:example} shows the example of a linked list node and a function
for inserting a new node after a given one. The {\tt TOID} declaration wraps the
linked list node type to produce a fat pointer to it. A {\tt TOID(ll\_node)}
type therefore is a pointer to a linked list node stored in persistent memory.
The {\tt insert\_after} function uses the {\tt D\_RO} and {\tt D\_RW} calls to
convert fat pointers into directly accessible read-only and read-write pointers
respectively. Other than the mechanics of using fat pointers the code entirely
resembles that which a programmer might write for a standard linked list in
DRAM.

\subsection{Failure Atomicity}
Programmers typically find it hard to reason about failures when writing
code. This is a problem when directly manipulating data structures in persistent
memory. Depending on the specifics of the system architecture, updates to
persistent memory can be persisted out of order thereby leaving the data
structure in an inconsistent state. To illustrate this consider a failure during
execution of the code to add a node to a linked list in
Figure~\ref{fig:example}. If either of the two updates to the next fields of the
previous and new node fail to be persisted, the linked list is left in a
disconnected state with one or more connected nodes inaccessible.

NVML provides tools for programmers to write code that is failure
atomic by allowing a block of code to be delimited as a transaction. A
transaction is either completely executed in terms of its effects on
persistent memory or not executed at all. Under the hood, NVML tracks all
updates to persistent memory in a persistent undo log - the log entries are
persisted before the update to the actual memory is allowed. On a failure, the
undo log is used to roll back incomplete transactions. A transaction is
considered complete when it completes execution and the updated objects are
flushed (automatically by NVML) to persistent memory. The log entries for the
transaction are then immediately deleted. The only burden placed on a programmer
for failure atomicity is to delimit transactions and that they must inform NVML
about objects that are updated. Figure~\ref{fig:example_fatomic} shows how to
convert the function for adding a node to a linked list into a failure atomic
one. The only additions are the transaction delimiters and the {\tt TX\_ADD}
calls to mark the objects being updated \emph{before} the actual updates to
them.

\begin{figure}
{ \scriptsize
\begin{verbatim}

struct ll_node {
  int value;
  TOID(ll_node) next;
};

@@

\end{verbatim}
}
\caption{Failure Atomic Peristent Linked List}
\label{fig:example_fatomic}
\end{figure}

A crucial feature of NVML is that it abstracts the details of the persistence
mechanism away from programmers. This allows the same code to be deployed on
across a range of persistence mechanisms - mapping pages from secondary
storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
save DRAM state to SSD on failure [FARM] or using directly accessed persistent
memory [PMFS]. Each of these options has different requirements. For example the
NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
instructions. NVDIMMs require only a CLFLUSH to flush the
cacheline and ensure that the line enters the asynchronous data refresh
domain. Using secondary storage is also an option with faster 3DXPoint storage
behind an NVMe interface - this would require an {\tt msync} and OS pagecache
intervention to persist data to secondary storage. NVML seamlessly uses the
right persistence mechanism without requiring code using the library to be aware
of these details.

\section{Replication}
Cyclone is designed to add fault tolerance to NVML client server applications
via replication - a quorum of machines maintains \emph{equivalent} state. Server
state is queried and manipulated via RPC calls from clients. Cyclone's approach
to replication is simple - we replicate the RPC call itself across replicas
rather than replicating every access made during execution of the RPC call,
which would be prohibitively expensive.

Cyclone provides strongly consistent replication - every machines in the quorum
of replicas maintains a log of RPC calls. All RPC calls (including read-only
calls that do not change state and therefore do not need replication) are always
executed by the leader replica. Cyclone treats the RPC call itself as a variable
sized opaque blob of data, leaving the application free to choose how it wishes
to marshall call related details and argument. All machines agree on the
committed sequence of RPC calls in the log. We achieve this by running an
instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
different machines in synchronization. The standard replicated state machine
approach requires that commands (entries in the log) be committed before they
can be applied to the state machine i.e. in our case - the RPC call be
executed. However, the fact that RPC calls that modify state must themselves be
executed in a crash consistent transaction allows us to overlap execution with
replication (described in the next section).

\subsection{Decoupled Replication}
\label{sec:decouple}
Cyclone decouples execution of the RPC call on any replica from its replication
by performing them in parallel to save time. Normally, this is a problem with
any consenus protocol including RAFT as replication might failt for a number of
reasons primarily due to a change of leader and subsequent of rollback of
previously appended (but not committed) log entries. In order to achieve
decoupled replication therefore Cyclone depends on the ability of NVML to roll
back crash consistent transactions. The ability to roll back a crash consistent
transaction is important to tolerate power failure as incomplete transactions
must be rolled back after recovery from a power failure to preserve the
integrity of the heap. NVML also allows the user to roll back a transaction if
needed. Cyclone exploits this feature by delaying the commit of the failure
atomic transaction after execution till replication is complete. In the event
that replication fails, the transaction is aborted and any changes to the non
volatile heap are rolled back.

A pseudocode description of how the runtime handles decoupled replication is
shown in Figure~\ref{fig:async_rep}.

\begin{figure}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
TX_BEGIN {
  Execute RPC call and modify local NVM
  Block till replication result known
  if(replication failed) {
    TX_ABORT;
  }
} TX_END
\end{verbatim}
}
\caption{Asynchronous Replication}
\label{fig:async_rep}
\end{figure}

Asynchronous replication provides good performance and we believe would be the
common case for RPC calls replicated via Cyclone. However the program might on
occasion require successful replication to be a precondition to exection. A case
where we find this necessary is output commit - where the RPC call has external
side effects, such as communicating with other machines in the cluster and
therefore crash consistency is insufficient to roll back the execution of the
RPC call in the event that replication fails. We therefore provide users the
ability to declare at the client side that an RPC call should execute
synchronously at the server side. This results in Cyclone ensuring that the RPC
call is successfully replicated before it attempts execution on any of the
replicas.

\subsection{Network Stack}
Replication protocols are expensive to operate - fact that has been underlined
in multiple pieces of research [FARM, in a box]. It would therefore appear that
replication would become the bottleneck in Cyclone. We observed however that
previous work did not implement replication around a high performance network
stack possibly explaining why they chose both high performance hardware (40
GigE+ NICS, custom FPGA designs) as well as specific software APIs (one sided
RDMA with metadata co-location, application compiled down to the FPGA) that
bypass the network stack. As a counterpoint, we designed Cyclone around DPDK
[dpdk] a high performance user-space network stack. DPDK provides convenient
access to packets arriving at an ethernet NIC by moving them into packet buffers
with the contained cache lines also directly placed into the last level cache of
the CPU during DMA.

Cyclone separates the available cores on a machine into cores dedicated to work
related to the consensus protocol and cores decidated to running application
code. The division is configurable. Each RAFT core runs an \emph{independent}
instance of the RAFT protocol (Section~\ref{sec:ind}). RAFT cores pass work to
the application cores through lock-free double ended queues. A RAFT core
dispatches an incoming RPC call for execution simultaneously with sending it out
for replication. As explained in Section~\ref{sec:decouple} this is beneficial
for reducing latency to completion.

@@

\subsection{Zero Copy Replication}
Consensus protcols such as RAFT usually have a simple common case path. Once a
quorum leader is elected, the leader's primary job is to transmit new entries
appended to its persistent log to its followers. The primary job of a follower
is to receive these entries, append them to its persistent log and send
acknowledgments to the leader. An entry is considered committed when the leader
receives an acknowledgment for an entry from a majority of the quorum. Followers
receive intimation of committed entries as piggybacked information on subsequent
replication messages.

In order to avoid copying the incoming RPC call to a persistent memory range,
Cyclone uses persistent memory on the system itself to allocate space for packet
buffers. However although a packet is accessible at the core the most up to date
data for the memory region occupied by the packet is in the CPU cache and not
necessarily in memory. To ensure that packet data has truly been made
persistent, Cyclone issues a series of cacheline flushes. This ensures that the
packet data is persistent. Cyclone maintains its log as a circular sequence of
pointers to packet buffers and therefore the append operation is completed by
adding a pointer to the received packet and flushing the cacheline containing
the pointer.

Our aim when designing Cyclone was to be relatively insensitive to the size of
the RPC call - i.e. the size of the RAFT log entry being replicated. Since
Cyclone is targeted towards general purpose applications we intended to support
a range of RPC call sizes without a degradation in performance caused by copying
the RPC call contents for transmission to each follower replica. Hence, we built
a zero copy mechanism for the common case of a leader replica sending the same
log entry to a set of follower replicas. To achieve our zero copy goal, we
exploit the fact that DPDKs packet buffers allow prepending data without
physically moving the data in the packet buffer. Fig [figures/packets.bmp] shows
how the layout of a packet buffer changes at the leader during replication. The
received packets contains only the RPC call from the client. We prepend a RAFT
related header (containing the previous log index, previous log term, leader's
term and leader's commit point) to packet buffer itself. To this RAFT header we
then prepend an IPv4 header. Although we use the raw packet interface from DPDK
and do not use IP based routing, we set IP addresses in a dummy header to use
5-tuple steering in order to place the packet in the right queue at the
receiving NIC. This composite packet buffer is then sent to every replica. In
order to properly route the packet through switches, we allocate a small buffer
containing an ethernet header for each follower replica and \emph{chain} it in
front of the prepared packet - ensuring we do not need to change the contents
for every replica.

\subsection{Batching}
Although the common case path in RAFT is straightforward, it still adds many
hundreds of cycles for each log entry to be processed. This adds up to a few
microseconds, a significant amount in the face of sub 10 us network latency
between nodes with DPDK. We amortize this overhead when operating under load by
batching multiple RPC calls together and treating the batch as a single RAFT log
entry.  DPDK by itself can efficiently batch incoming packets under load -
allowing the user to easily receive a burst (currently limited to 32) of packets
rather than a single one for a polling call. DPDK further allows chaining packet
buffers together and treating the chain as a single packet. This is convenient
for our zero copy approach and we chain packet mbufs together to form a single
packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
single packet when being replicated, the length of a chain is limited by the
size of the coalesced packet, that must fit into an ethernet jumbo frame of
approximately 9K bytes.

\subsection{Multiple RAFT Instances}
\label{sec:ind}
We scale replication performance in Cyclone by using multiple cores on the same
machine to each run an independent instance of the RAFT protocol. For a quorum
of homogeneous machines, we dedicate the same core and NIC queue pair on each
machine to the same instance of RAFT. Each instance maintains an
\emph{independent} log. All the RAFT instances on a machine share the common set
of application cores, dispatching work to them via the (shared) lock-free double
ended queues (implemented as circular buffers in DPDK). It is important to note
that there is no fixed ordering between RPC calls dispatched for execution from
different RAFT instances. We discuss in Sections
~\ref{sec:exec},\ref{sec:example} how a Cyclone user can still impose
determinism for linearizable data structures, while continuing to scale across
the multiple RAFT instances.

Other than the lock free queues for dispatching work to application cores, the
multiple RAFT instances share nothing, being assigned dedicated NIC queue pairs
and hence also dedicated packet buffer pools associated with those queue
pairs. This removes any synchronization bottlenecks on memory allocation. We
note that DPDK uses reference counting to decide when to free a packet
buffer. RAFT cores continue to hold a reference to a packet containing an RPC
call even after dispatching it for execution until it has completed replicating
the contents of the packet and is ready to compact away that entry from its RAFT
log.

\subsection{RPC Call Execution}
\label{sec:exec}
We now describe how RPC calls are executed and results returned to the
client. There are two important aspects of RPC call execution in Cyclone -
steering, and at most once semantics.

An RPC call via Cyclone must traverse a specific RAFT instance (RAFT log) and
specific execution core. These details are provided by the user when making the
RPC call to the client side library provided by Cyclone. The user therefore is
aware of the number of execution cores and RAFT instances and given complete
control of \emph{RPC call steering}.

Cyclone provides at most once semantics. It supports a large but fixed (compile
time configured) number of clients. Each RPC call from a client bears a sequence
number. \emph{Each execution core independently enforces execution of RPC calls
  in increasing sequence numbers}. This is done by storing the last executed
sequence number in a per-core durable datastructure as part of the crash
consistent transaction wrapping the execution of the RPC call itself. Even in
the presence of fail-overs, this durable memory of sequence numbers seen from a
client ensures at most once semantics. In addition to guarding agains repeated
execution of the same call, Cyclone also remembers the \emph{results} of the
last executed RPC call. In order to assist clients (particularly stateless
clients), Cyclone allows the client to query the last seen sequence number and
result for that RPC call from the quorum of servers.

\section{Putting it together - A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practice by putting together an
example of a linearizable hash table - a useful and common data structure
[memcached]. The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys. Such a hash table is simple to derive from a
volatile hash table and readily available in NVML [nvml.io] as an example [src
  code link]. To use that (non client-server) example with Cyclone we add a
simple RPC call format, manually marshall and demarshall the (at most two)
arguments to the RPC call in an RPC call handler we wrote (a few tens of lines
of code) and wrote a simple client driver program (a few tens of additional
lines of code).

@@

There are two challenges in scaling the performance and improving the utility of
the linearizable and durable hash table. The first is to ensure linearizability
while scaling performance by adding RAFT cores and application cores. The second
is to support distributed transactions - that manipulate a set of key-value
pairs atomically in the face of failure.

@@

\subsection{Scaling}
Scaling the performance of the hash table by using multiple RAFT cores and
execution cores presents determinism related challenges. Two RPC calls being
steered through different RAFT cores can end up executing in different orders on
different replicas even if steered to the same execution core. At the same time,
two RPC calls steered through the same RAFT core but dispatched to different
execution cores can end up completing in different orders in different replicas.

Our solution to this problem in Cyclone is one that applies to most durable
data structures - not just the hash table we consider here. We observed that
replicas of the durable data structure require \emph{per-key determinism}
and not determinism across \emph{all} operations. We therefore steer
operations to the same key to the same RAFT core and the same execution core, by
hashing the key to determine the RAFT core and execution core. The actual
configuration of the hash table therefore can be different across different
replicas - for example due to the insertion of two keys mapping to the same hash
bucket completing in different orders across different replicas. However, since the
results of insert, delete and lookup operations for a key depend only on
previous operations to the same key the replicas provide identical results for
committed operations.

@@

\subsection{Distributed Transactions}
A useful primitive for many distributed systems is the ability to run
distributed transactions - an atomic sequence of conditional updates spanning
multiple machines. Distributed transactions have two distinct facets:
atomicity and fault tolerance. Distributed transactions appear atomic with
respect to other atomic transactions and simple updates or reads by virtue of
standard concurrency control techniques. The most common technique is two phase
locking: cquire locks, verify that previously read objects are at the desired
values, update the locked objects and finally release the locks. A survey of
various distributed systems and informal discussions with programmers led us to
the conclusion that this aspect of distributed transactions was fairly easy to
understand and implement. The hard part for most programmers is the second
aspect: fault tolerance. A distributed transaction spans many machines and has a
large footprint to clean up in the event of failure. Failure handling is made
more complex when considering cascading failures - further failures while
recovering from a failure. Cyclone was designed to allow primitives such as
distributed transctions to be added without the programmer needing to worry
about faults or write fault recovery code - as we demonstrate with the hash
table.

We support distributed transactions in our hash table by allowing the user to
specify a group of insert/delete/lookup operations that must occur
atomically. We use the existing bucket locks to implement a simple two phase
locking protocol to implement the transction. A client makes an RPC call to the
server specifying the transaction and the server executes it on the client's
behalf. Figure [] shows pseudocode for the transaction. Execution of the
transaction has four distinct phases: acquire locks, verify values, apply
updates and finally release locks.

@@

Next, we need to ensure that although every machine in the quorum makes all
necessary RPC calls, the call itself is executed only once. This is achieved by
using Cyclone's at most once semantics. All machines in the quorum executing the
transaction use the same client identifier and the same montonically increasing
sequence of RPC call numbers. Finally, we need to deal with the fact that
Cyclone only remembers the last executed RPC call for a client. This means that
if different machines in the quorum move at different speeds, some machines will
not receive a response to the RPC call - they will instead receive an error that
tells them that they are too far behind. To ensure that the transaction is still
guaranteed to complete, we use a \emph{different} client identifier for each
phase. Thus, until the transaction completes and a new one is started the
response for any RPC call continues to be available. Figure [] details in
pseudocode the execution of a transaction with these added details. Crucially
there is no need for awareness of faults nor any need to write fault recovery
code. Finally, we need to ensure that the client can reliably query the result
of the transaction. If the leader of the quorum executing the transaction falls
behind it cannot supply the result. We therefore immediately return and
acknowledment to the client as soon as the RPC call for the transaction has been
replicated. We store the result of the transaction in a special data structure
updated through an additional RPC call made at the end of the transaction. The
client queries this data structure to determine the fate of the transaction.

\section{Evaluation}
\begin{figure}
  Source: results_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
  ports = 4. Fix payload = 0. One line per number of replicas (1,2,3).
  \caption{Replicaton Performance - Varying replica count}
  \label{fig:perf_rep}
\end{figure}

\begin{figure}
  Source: results_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
  replicas = 3. Fix payload = 0. One line per number of ports.
  \caption{Replicaton Performance - Varying ethernet ports}
  \label{fig:perf_ports}
\end{figure}

>>>>

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig,endnotes}
\begin{document}

%don't want date printed
\date{}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf Cyclone: Replication Middleware for NVM Applications}

%for single author (just remove % characters)
%\author{
%  {\rm Paper #: XXX }\\
%  XXX
%  \and
%  {\rm YYY}\\
%  YYY
%} % end author

\maketitle

% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
\thispagestyle{empty}

\subsection*{Abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. In anticipation, libraries that allow programmers to build durable
data structures on non-volatile heaps have become available. A missing piece
however is a component to make such durable data structures available using
replication across a commodity network such as ethernet. Existing solutions that
offer good performance are biased towards pure transactions across RDMA with a
dependence on external services and application specific code for fault recovery.

Cyclone is replication middleware that aims to add fault tolerance as
transparently as possible. A design for Cyclone is that NVM application
developers should not need to add fault recovery code. A second design goal for
Cyclone is to strive for good performance on commodity ethernet - with
innovations in the software stack rather than a requirement for RDMA capable
networking hardware.

Cyclone achieves these goals by providing a simple API for client server
applications that allows server state to be replicated by executing the same
sequence of RPC calls across all replicas. It provides at most once semantics
with automatic failover to hide fault recovery from programmers. Finally it uses
DPDK - a high performance, low latency network stack coupled with batching,
zero-copy and the ability to scale replication performance across
multiple instances of a consensus protocol to replicate millions of small
payload RPC messages a second on commodity ethernet.

\section{Introduction}
TBD
\section{Single Node Programming Model}
Cyclone starts with the assumption that the programmer has already built a
concurrent durable application on a non-volatile heap. A number of libraries and
filesystems make this possible today [DAX, NVML, others ?]. As a starting point,
we support applications built on top of Intel's Non Volatile Memory Library
(NVML) [nvml.io]. We describe NVML's programming model in this section to provide
adequate background for the rest of this paper. The primary feature of interest
is the ability to write code that is crash consistent. In other words code that
makes changes to the non-volatile heap is enclosed in a crash consistent
transaction such that either all of the transaction executes or none of it
executes in the face of power failures. This is critical to maintaining
consistency as the data structure, being durable, must survive across power
failures. To illustrate this idea, Figure~\ref{fig:example} shows how one might
build a persistent linked list using NVML - the key idea is to enclose the
entire process of updating the linked list within a crash consistent
transaction. This ensure that the linked list is not corrupted due to a power
failure. A second point worth noting is that NVM libraries such as NVML usually
provide a way to wrap pointers to non-volatile memory in order that such memory
may be mapped to different locations in virtual memory across restarts without
impact to the accessing application.

\begin{figure}
  { \scriptsize
\begin{verbatim}
struct ll_node {
  int value;
  TOID(ll_node) next;
};

@@

\end{verbatim}
  }
\caption{A Persistent Linked List}
\label{fig:example}
\end{figure}

It is easy to turn the linked list example in NVML into a concurrent linked list
by simply adding a lock to it. Such a lock can be maintined in volatile DRAM if
so desired for performance. More sophisticated fine-grained or optimistic
concurrency control mechanisms for NVM are also possible [mnemosyne].

\section{Replication}
Cyclone is designed to add fault tolerance to NVML client server applications
via replication - a quorum of machines maintains \emph{equivalent} state. Server
state is queried and manipulated via RPC calls from clients. Cyclone replicates
the RPC call itself across replicas rather than replicating every access made
during execution of the RPC call.

Cyclone provides strongly consistent replication - every machine in the quorum
of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as a variable
sized opaque blob of data, leaving the application free to choose how it wishes
to marshall call related details and argument. All machines agree on the
committed sequence of RPC calls in the log. We achieve this by running an
instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
different machines in synchronization. Specially flagged `read-only` RPC calls
are executed on the leader replica and are not replicated.

\subsection{Decoupled Replication}
\label{sec:decouple}
The standard replicated state machine approach requires that commands (entries
in the log) be committed before they can be applied to the state machine i.e. in
our case - the RPC call be executed. In contrast, Cyclone decouples execution of
the RPC call on any replica from its replication.

However, the fact that RPC calls that modify state must themselves be executed
in a crash consistent transaction allows us to overlap execution with
replication. Normally, this would be a problem as replication might failt for a
number of reasons but primarily due to a change of leader and subsequent of
rollback of previously appended (but not committed) log entries. To solve this
problem, Cyclone delays the commit of the failure atomic transaction after
execution till replication is complete. In the event that replication fails, the
transaction is aborted and any changes to the non volatile heap are rolled back.

A pseudocode description of how the runtime handles \emph{decoupled replication}
is shown in Figure~\ref{fig:async_rep}.

\begin{figure}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments 
TX_BEGIN { 
  Execute RPC call and modify local NVM 
  Block till replication result known 
  if(replication failed) {
     TX_ABORT 
  } 
} TX_END
\end{verbatim}
}
\caption{Decoupled Replication}
\label{fig:async_rep}
\end{figure}

Asynchronous replication provides good performance and we believe it would be the
common case for RPC calls replicated via Cyclone. However the program might on
occasion require successful replication to be a precondition to exection. A case
where we find this necessary is output commit - where the RPC call has external
side effects, such as communicating with other machines in the cluster and
therefore crash consistency is insufficient to roll back the execution of the
RPC call in the event that replication fails. We therefore provide users the
ability to declare at the client side that an RPC call should execute
synchronously at the server side. This results in Cyclone ensuring that the RPC
call is successfully replicated before it attempts execution on any of the
replicas.

\subsection{Network Stack}
Cyclone is designed around DPDK [dpdk] a high performance network stack. DPDK
provides convenient access to packets arriving at an ethernet NIC by moving them
into packet buffers with the contained cache lines also directly placed into the
last level cache of the CPU during DMA [ddio].

Cyclone separates the available CPU cores on a machine, in a configurable way,
into cores dedicated to work related to the consensus protocol and cores
decidated to running application code. Each RAFT core runs an \emph{independent}
instance of the RAFT protocol. RAFT cores pass work to the application cores
through lock-free double ended queues. A RAFT core dispatches an incoming RPC
call for execution simultaneously with sending it out for replication.

@@

\subsection{Zero Copy Replication}
Consensus protcols such as RAFT usually have a simple common case path. Once a
quorum leader is elected, the leader's primary job is to transmit new entries
appended to its persistent log to its followers. The primary job of a follower
is to receive these entries, append them to its persistent log and send
acknowledgments to the leader. An entry is considered committed when the leader
receives an acknowledgment for an entry from a majority of the quorum. Followers
receive intimation of committed entries as piggybacked information on subsequent
replication messages. The common case path at the leader lends itself to zero
copy since the same set of log entries is sent to all replicas.

In order to avoid copying the incoming RPC call contents to a persistent memory
range, Cyclone uses persistent memory on the system itself to allocate space for
packet buffers. However although a packet is accessible at the core the most up
to date data for the memory region occupied by the packet is in the CPU cache
and not necessarily in memory. To ensure that packet data has truly been made
persistent, Cyclone issues a series of cacheline flushes. This ensures that the
packet data is persistent. Cyclone maintains its log as a circular sequence of
pointers to packet buffers and therefore the append operation is completed by
adding a pointer to the received packet and flushing the cacheline containing
the pointer.

We exploit the fact that DPDKs packet buffers allow prepending data without
physically moving the data in the packet buffer. Fig [figures/packets.bmp] shows
how the layout of a packet buffer changes at the leader during replication. The
received packets contains only the RPC call from the client. We prepend a RAFT
related header (containing the previous log index, previous log term, leader's
term and leader's commit point) to packet buffer itself. To this RAFT header we
then prepend an IPv4 header. We set IP addresses in the header to use
5-tuple steering in order to place the packet in the right queue at the
receiving NIC. This composite packet buffer is then sent to every replica. In
order to properly route the packet through switches, we allocate a small buffer
containing an ethernet header for each follower replica and \emph{chain} it in
front of the prepared packet - ensuring we do not need to change the contents
of the packet itself for every replica.

\subsection{Batching}
Although the common case path in RAFT is straightforward, it still adds many
hundreds of cycles for each log entry to be processed. This adds up to a few
microseconds, a significant amount in the face of sub 10 us network latency
between nodes with DPDK. We amortize this overhead when operating under load by
batching multiple RPC calls together and treating the batch as a single RAFT log
entry.  DPDK by itself can efficiently batch incoming packets under load -
allowing the user to easily receive a burst (currently limited to 32) of packets
rather than a single one for a polling call. DPDK further allows chaining packet
buffers together and treating the chain as a single packet. This is convenient
for our zero copy approach and we chain packet mbufs together to form a single
packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
single packet when being replicated, the length of a chain is limited by the
size of the coalesced packet, that must fit into an ethernet jumbo frame of
approximately 9KB.

\subsection{RPC Call Execution}
\label{sec:exec}
We now describe how RPC calls are executed and results returned to the
client. There are two important aspects of RPC call execution in Cyclone -
steering and at most once semantics.

An RPC call via Cyclone must traverse a specific RAFT core (and associated
consensus protocol instance) and specific application core. These details are
provided by the user when making the RPC call to the client side library
provided by Cyclone. The user therefore is aware of the number of execution
cores and RAFT instances and given complete control of \emph{RPC call steering}.
For most applications we envisage a simple modulo hashing scheme should suffice
for load balancing.

Cyclone provides at most once semantics. It supports a large but fixed (at
startup time) number of clients. Each RPC call from a client bears the client
number and a sequence number. \emph{Each execution core independently enforces
  execution of RPC calls in increasing sequence numbers}. This is done by
storing the last executed sequence number in a per-core durable datastructure as
part of the crash consistent transaction wrapping the execution of the RPC call
itself. Even in the presence of fail-overs, this durable memory of sequence
numbers seen from a client ensures at most once semantics. In addition to
guarding agains repeated execution of the same call, Cyclone also remembers the
\emph{results} of the last executed RPC call. This is done to assist clients
(particularly stateless clients). Cyclone allows the client to query the last
seen sequence number and result for that RPC call from the quorum of servers.

\section{A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practice by putting together an
example of a linearizable hash table - a useful and common data structure.
[memcached]. The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys. Such a hash table is simple to derive from a
volatile hash table and readily available in NVML [nvml.io] as an example [src
  code link]. To use that (non client-server) example with Cyclone we added a
simple RPC call format, manually marshalled and demarshalled the few
arguments to the RPC call in an RPC call handler we wrote - a few tens of lines
of code. Finally, we added a simple client driver program (a few tens of
additional lines of code).

@@

There are two interesting challenges in scaling the performance and improving
the utility of the linearizable and durable hash table. The first is to ensure
linearizability while scaling performance by adding RAFT cores and application
cores. The second is to support distributed transactions - that manipulate a set
of key-value pairs atomically in the face of failure.

@@

\subsection{Scaling}
Scaling the performance of the hash table by using multiple RAFT cores and
application cores presents determinism related challenges. Two RPC calls being
steered through different RAFT cores can end up executing in different orders on
different replicas even if steered to the same application core. At the same time,
two RPC calls steered through the same RAFT core but dispatched to different
application cores can end up completing in different orders on different replicas.

Our solution to this problem in Cyclone is to observe that replicas of the hash
table require \emph{per-key determinism} and not determinism across \emph{all}
operations. We therefore steer operations to the same key to the same RAFT core
and the same application core, by hashing the key to determine the RAFT core and
execution core. The actual configuration of the hash table therefore can be
different across different replicas - for example due to the insertion of two
keys mapping to the same hash bucket completing in different orders across
different replicas. However, since the results of insert, delete and lookup
operations for a key depend only on previous operations to the same key the
replicas provide identical results for committed operations.

@@

\subsection{Distributed Transactions}
A useful primitive for many distributed systems is the ability to run
distributed transactions - an atomic sequence of conditional updates spanning
multiple machines. Distributed transactions have two distinct facets: atomicity
and fault tolerance. Distributed transactions appear atomic with respect to
other atomic transactions and simple updates or reads by virtue of standard
concurrency control techniques. The most common technique is two phase locking:
cquire locks, verify that previously read objects are at the desired values,
update the locked objects and finally release the locks. We believe that NVM
application programmers - particularly those having experience with concurrent
code should not find it hard to deal with distributed concurrency. The hard part
for most programmers is likely to be the second aspect: fault tolerance. A
distributed transaction spans many machines and has a large footprint to clean
up in the event of failure. Failure handling is made more complex when
considering cascading failures - further failures while recovering from a
failure. Cyclone was designed to allow primitives such as distributed
transctions to be added without the programmer needing to worry about faults or
write fault recovery code. As an exercise we show how to add support for
distributed transactions to our hash table.

We support distributed transactions in our hash table by allowing the user to
specify a group of insert/delete/lookup operations that must occur
atomically. We use the existing bucket locks to implement a simple two phase
locking protocol to implement the transction. A client makes an RPC call to the
server specifying the transaction and the server executes it on the client's
behalf. Execution of the transaction has four distinct phases: acquire locks,
verify values, apply updates and finally release locks.

@@

Next, we need to ensure that although every machine in the quorum makes all
necessary RPC calls, the call itself is executed only once. This is achieved by
using Cyclone's at most once semantics. All machines in the quorum executing the
transaction use the same client identifier and the same montonically increasing
sequence of RPC call numbers. Finally, we need to deal with the fact that
Cyclone only remembers the last executed RPC call for a client. This means that
if different machines in the quorum move at different speeds, some machines will
not receive a response to the RPC call - they will instead receive an error that
tells them that they are too far behind.

To ensure that the transaction is still guaranteed to complete, we use a
\emph{different} client identifier for each phase. Thus, until the transaction
completes and a new one is started the response for any RPC call continues to be
available. Figure [] shows in pseudocode the server side details of a
transaction. It is worth comparing the lack of any need for explicit fault
recovery in Cyclone with fault recovery in systems systems that explicity
require such cleanup [FARM, also refer to pages].  We believe that making
availability easy for NVM applications is key to increasing the adoption rate of
NVM by data center operators.

\section{Evaluation}
\begin{figure}
  Source: results\_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
  ports = 4. Fix payload = 0. One line per number of replicas (1,2,3).
  \caption{Replicaton Performance - Varying replica count}
  \label{fig:perf_rep}
\end{figure}

\begin{figure}
  Source: results\_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
  replicas = 3. Fix payload = 0. One line per number of ports.
  \caption{Replicaton Performance - Varying ethernet ports}
  \label{fig:perf_ports}
\end{figure}
